---
title: "Data for National Water Use Model"
author: "Scott Worland"
date: "November 22 2016"
output: pdf_document
header-includes:
  - \usepackage{longtable}
---
  
```{r, echo=F}
knitr::opts_chunk$set(fig.width=7,fig.height=5,fig.align = 'center', warning=F, message=F,cache=F)
knitr::opts_knit$set(root.dir = '~/Water Conservation/R_conservation/USGSwaterUse/usgs-wu-mlm/Fall_2016')

library(pacman)
pacman::p_load(ggplot2,dplyr,reshape2,xtable,magrittr,
               stringr,readxl,maps,choroplethr,ggmap,purrr, 
               viridis,acs,feather,gridExtra,corrr)

setwd("~/Water Conservation/R_conservation/USGSwaterUse/usgs-wu-mlm/Fall_2016")
```

```{r, echo=F}
# mask read_excel to surpress output
read_excel <-  function(...) {
  quiet_read <- purrr::quietly(readxl::read_excel)
  out <- quiet_read(...)
  if(length(c(out[["warnings"]], out[["messages"]])) == 0)
    return(out[["result"]])
  else readxl::read_excel(...)
}

summarize = dplyr::summarize
rename = dplyr::rename
```

# Response variable: county level water use

The first step is to [load the data](http://water.usgs.gov/watuse/data/2010/index.html). A data dictionary for the USGS data can be found [here](http://water.usgs.gov/watuse/data/2010/datadict.html).

```{r}
# build state name and abb data.frame
states <- data.frame(state.abb,state=state.name, stringsAsFactors = F) %>%
  bind_rows(data.frame(state.abb="DC", state="district of columbia", 
                       stringsAsFactors = F))

# url for wudata file
wudata.file <- "http://water.usgs.gov/watuse/data/2010/usco2010.txt"

# load datafile and add columns
wudata <- read.delim(wudata.file, stringsAsFactors=F) %>%
  setNames(str_to_lower(names(.))) %>%
  filter(!state %in% c("AK","HI","PR","VI")) %>%
  select(cnty_fips=fips, state.abb=state, county, year, cnty_pop=tp.totpop,
         pop_served=ps.topop, wthdrw_gw=ps.wgwfr, wthdrw_sw=ps.wswfr, 
         wthdrw_tw=ps.wfrto) %>%
  left_join(states, by="state.abb") %>%
  mutate(cnty_fips = str_pad(cnty_fips,5, pad = "0"),# add leading zero
         county = str_replace_all(county, fixed(" County"), ""), # remove word "County"
         cnty_pop = cnty_pop*1000, # convert population to raw pop
         pop_served = pop_served*1000,
         prop_sw=wthdrw_sw/wthdrw_tw)
```


# Grouping variables

## County typology codes

The USDA publishes a [county typology data set](http://www.ers.usda.gov/data-products/county-typology-codes.aspx). From the webpage: "The 2015 County Typology Codes classify all U.S. counties according to six mutually exclusive categories of economic dependence and six overlapping categories of policy-relevant themes. The economic dependence types include farming, mining, manufacturing, Federal/State government, recreation, and nonspecialized counties. The policy-relevant types include low education, low employment, persistent poverty, persistent child poverty, population loss, and retirement destination." Description codes can be found [here](https://wayback.archive-it.org/5923/20110913215735/http://www.ers.usda.gov/Briefing/Rurality/RuralUrbCon/).

```{r, eval=F}
# url for typology data
typ.url <- "http://www.ers.usda.gov/webdocs/DataFiles/County_Typology_Codes__17972//
all_final_codes.xls"

# download typology file
download.file(typ.url, "data/cnty_typology_2004.xls")

# load typology file into R
cnty_typology <- read_excel("data/cnty_typology_2004.xls", sheet = 2, col_names = T) %>%
  select(cnty_fips=FIPSTXT,rur2urb=rururb2003,econdep,retire) %>%
  mutate_each(funs(as.character))

# add more meaningful descriptions
## urban continuum
rur2urb_des <- data.frame(
  rur2urb = as.character(c(1:9)),
  rur2urb_des = c("metro_large", # pop >= 1e6
                  "metro_med", # pop 250,000 to 1e6
                  "metro_small", # pop < 250,000
                  "urb_large_adj", # pop >= 20,000, metro adjacent 
                  "urb_large_det", # pop >= 20,000, metro detached
                  "urb_small_adj", # pop 2500-20,000, metro adjacent
                  "urb_small_det", # pop 2500-20,000, metro detached
                  "rural_adj", # pop < 2500, metro adjacent
                  "rural_det"), # pop < 2500, metro detached
  stringsAsFactors = F)

## economic dependencies
econdep_des <- data.frame(
  econdep = as.character(c(1:6)),
  econdep_des = c("farming", # dependent on farming
                  "mining", # dependent on mining
                  "manufacuring", # dependent on manufactoring
                  "government", # dependent on federal and/or state govt
                  "services", # dependent on services
                  "nonspecialized"), # nonspecialized
  stringsAsFactors = F)

cnty_typology %<>% 
  left_join(rur2urb_des, by="rur2urb") %>% 
  left_join(econdep_des, by="econdep") %>%
  select(cnty_fips, rur2urb=rur2urb_des, econdep=econdep_des,retire)
```

```{r, echo=F}
# load typology file into R
cnty_typology <- read_excel("data/cnty_typology_2004.xls", sheet = 2, col_names = T) %>%
  select(cnty_fips=FIPSTXT,rur2urb=rururb2003,econdep,retire) %>%
  mutate(cnty_fips=as.numeric(cnty_fips) %>%
           round(.,0) %>%
           as.character() %>%
           str_pad(5, pad = "0")) %>%
  mutate_each(funs(as.character))

# add more meaningful descriptions
## urban continuum
rur2urb_des <- data.frame(
  rur2urb = as.character(c(1:9)),
  rur2urb_des = c("metro_large", # pop >= 1e6
                  "metro_med", # pop 250,000 to 1e6
                  "metro_small", # pop < 250,000
                  "urb_large_adj", # pop >= 20,000, metro adjacent 
                  "urb_large_det", # pop >= 20,000, metro detached
                  "urb_small_adj", # pop 2500-20,000, metro adjacent
                  "urb_small_det", # pop 2500-20,000, metro detached
                  "rural_adj", # pop < 2500, metro adjacent
                  "rural_det"), # pop < 2500, metro detached
  stringsAsFactors = F)

## economic dependencies
econdep_des <- data.frame(
  econdep = as.character(c(1:6)),
  econdep_des = c("farming", # dependent on farming
                  "mining", # dependent on mining
                  "manufacuring", # dependent on manufactoring
                  "government", # dependent on federal and/or state govt
                  "services", # dependent on services
                  "nonspecialized"), # nonspecialized
  stringsAsFactors = F)

cnty_typology %<>% 
  left_join(rur2urb_des, by="rur2urb") %>%
  left_join(econdep_des, by="econdep") %>%
  select(cnty_fips, rur2urb=rur2urb_des, econdep=econdep_des,retire)
```

## Climate regions

NOAA list [9 climate regions](https://www.ncdc.noaa.gov/monitoring-references/maps/us-climate-regions.php) by state which might also serve as an interesting grouping variable. I was unable to find the actual data online so made a csv using the map linked above.

```{r}
## climate regions
climate_region <- read.csv("data/climate_regions.csv", header=T, stringsAsFactors = F) %>%
  rename(state.abb=state)
```

\newpage

# Explanatory Variables

## Temperature and precipitation

There are plenty of ways to aggregate gridded temp and precip data into counties. However, using the USGS `geoknife` [R package](https://cran.r-project.org/web/packages/geoknife/vignettes/geoknife.html) allows for a very reproducible approach. `geoknife` primarily works through remote servers so much of the work is not done on the actual machine.

### Create web data and clip to counties

```{r, eval=F}
summarize = dplyr::summarise
query = geoknife::query

# query available variables in prism dataset
prism = webdata(url='http://cida.usgs.gov/thredds/dodsC/prism')
query(prism,'variables')

# load data
precip <- webdata(list(
  times = as.POSIXct(c('1970-01-01','2010-12-01')),
  url = 'http://cida.usgs.gov/thredds/dodsC/prism',
  variables = 'ppt'))

tmax <- webdata(list(
  times = as.POSIXct(c('1970-01-01','2010-12-01')),
  url = 'http://cida.usgs.gov/thredds/dodsC/prism',
  variables = 'tmx'))

tmin <- webdata(list(
  times = as.POSIXct(c('1970-01-01','2010-12-01')),
  url = 'http://cida.usgs.gov/thredds/dodsC/prism',
  variables = 'tmn'))

# provide FIPS for every county
library(maps) #load fips from maps package
cnty.fips <- stringr::str_pad(as.character(maps::county.fips$fips), pad = '0', width = 5)

# create stencil
all.cnty <- webgeom(geom='sample:Counties', attribute='FIPS', values=cnty.fips)

# download data
precip.job <- geoknife(stencil=all.cnty, fabric=precip, wait = F, email = 'scworland@usgs.gov')
tmax.job <- geoknife(stencil=all.cnty, fabric=tmax, wait = F, email = 'scworland@usgs.gov')
tmin.job <- geoknife(stencil=all.cnty, fabric=tmin, wait = F, email = 'scworland@usgs.gov')

# build seasons data frame
seasons <- data.frame(month=c("Mar","Apr","May",
                              "Jun","Jul","Aug",
                              "Sep","Oct","Nov",
                              "Dec","Jan","Feb"),
                      season=c(rep("spring",3),
                               rep("summer",3),
                               rep("fall",3),
                               rep("winter",3)),
                      stringsAsFactors = F)

```

### extract precip data and calculate statistics

```{r,eval=F}
# extract precip data
cnty.precip <- result(precip.job) 

# mean summer precip 1970-2010
mean.precip <- cnty.precip %>%
  mutate(year=format(DateTime,'%Y'),
         month = months(DateTime, abbreviate=T)) %>% 
  select(-DateTime,-year,-statistic,-variable) %>%
  melt(., id.vars="month") %>%
  set_colnames(c("month","cnty_fips","ppt")) %>%
  mutate(cnty_fips = as.character(cnty_fips),
         ppt = as.numeric(ppt)) %>%
  left_join(seasons, by="month") %>%
  group_by(season,cnty_fips) %>%
  summarize(summer_mean40yr = mean(ppt, na.rm=T)) %>%
  filter(season=="summer") %>%
  ungroup()

# mean summer precip 2005-2010 minus 40 yr mean
precip05_10 <- cnty.precip %>%
  mutate(year=format(DateTime,'%Y'),
         month = months(DateTime, abbreviate=T)) %>% 
  select(-DateTime,-statistic,-variable) %>%
  melt(., id.vars=c("year","month")) %>%
  set_colnames(c("year","month","cnty_fips","ppt")) %>%
  mutate(cnty_fips = as.character(cnty_fips),
         ppt = as.numeric(ppt)) %>%
  filter(year %in% c(as.character(2005:2010))) %>%
  left_join(seasons, by="month") %>%
  group_by(season,cnty_fips) %>%
  summarize(summer_mean05_10 = mean(ppt, na.rm=T)) %>%
  filter(season=="summer") %>%
  ungroup() %>%
  select(-season) %>%
  left_join(mean.precip, by="cnty_fips") %>%
  mutate(ppt_diff=summer_mean05_10-summer_mean40yr) %>%
  ungroup()
```

```{r, echo=F}
## mean summer precip 1970-2010
precip05_10 <- read_feather("data/precip_diff_40yr.feather") %>%
  select(cnty_fips,ppt_0510=summer_mean05_10,
         ppt_40=summer_mean40yr,ppt_diff)

mean.precip.map <- precip05_10 %>%
  select(region=cnty_fips, value=ppt_diff) %>%
  mutate(region=as.integer(region),
         value=as.numeric(value))
  
county_choropleth(mean.precip.map , num_colors = 8, 
                  title="2005-2010 summer precip - 40 yr summer precip") +
  scale_fill_brewer(name="ppt diff", palette='YlGnBu', drop=FALSE)
```

### extract max temp data and calculate statistics

```{r, eval=F}
# extract temp data
cnty.temp <- result(tmax.job) 

# mean max summer temp 1970-2010
mean.tmax <- cnty.temp %>%
  mutate(year=format(DateTime,'%Y'),
         month = months(DateTime, abbreviate=T)) %>% 
  select(-DateTime,-year,-statistic,-variable) %>%
  melt(., id.vars="month") %>%
  set_colnames(c("month","cnty_fips","mxtmp")) %>%
  mutate(cnty_fips = as.character(cnty_fips),
         mxtmp = as.numeric(mxtmp)) %>%
  left_join(seasons, by="month") %>%
  group_by(season,cnty_fips) %>%
  summarize(summer_mean40yr = mean(mxtmp, na.rm=T)) %>%
  filter(season=="summer") %>%
  ungroup()

# mean max summer temp 2005-2010 minus 40 yr mean
tmax05_10 <- cnty.temp %>%
  mutate(year=format(DateTime,'%Y'),
         month = months(DateTime, abbreviate=T)) %>% 
  select(-DateTime,-statistic,-variable) %>%
  melt(., id.vars=c("year","month")) %>%
  set_colnames(c("year","month","cnty_fips","mxtmp")) %>%
  mutate(cnty_fips = as.character(cnty_fips),
         mxtmp = as.numeric(mxtmp)) %>%
  filter(year %in% c(as.character(2005:2010))) %>%
  left_join(seasons, by="month") %>%
  group_by(season,cnty_fips) %>%
  summarize(summer_mean05_10 = mean(mxtmp, na.rm=T)) %>%
  filter(season=="summer") %>%
  ungroup() %>%
  select(-season) %>%
  left_join(mean.tmax, by="cnty_fips") %>%
  mutate(tmax_diff=summer_mean05_10-summer_mean40yr) %>%
  ungroup()
```

```{r, echo=F}
## mean summer temp 1970-2010
tmax05_10 <- read_feather("data/tmax_diff_40yr.feather") %>%
  select(cnty_fips,tmax_0510=summer_mean05_10,
         tmax_40=summer_mean40yr,tmax_diff)

mean.tmax.map <- tmax05_10 %>%
  select(region=cnty_fips, value=tmax_diff) %>%
  mutate(region=as.integer(region),
         value=as.numeric(value))

county_choropleth(mean.tmax.map, num_colors = 8, 
                  title="2005-2010 summer max temp - 40 yr summer max temp") +
  scale_fill_brewer(name="tmax diff", palette='YlOrRd', drop=FALSE)
```

## Water Yield

Thomas Brown from Colorado State sent me the water yield data from [this](http://www.fs.fed.us/rmrs/documents-and-media/really-mean-annual-renewable-water-supply-contiguous-united-states) recent study.


"Daily water yield was estimated using the Variable Infiltration Capacity (VIC) model at each 1/8ยบ by 1/8ยบ (about 12 km by 12 km) grid cell across the conterminous U.S. Yields were aggregated over time to estimate mean annual yield. Having the spatially distributed estimates of mean annual water yield at its source, land boundaries were then overlaid. Aggregating estimates of yield across cells within a boundary indicates the amount of water supply originating within the designated area."

```{r}
## water yield
wyield <- read.csv("data/VIC_Q_by_CountyFIPS.csv", header=T, 
                   colClasses = c("character","numeric","numeric")) %>%
  mutate(cnty_fips=str_pad(cnty_fips, pad = '0', width = 5))
```

```{r, echo=F}
wyeild.map <- wyield %>%
  select(region=cnty_fips, value=Qmm) %>%
  mutate(region=as.integer(region),
         value=as.numeric(value))

county_choropleth(wyeild.map, num_colors = 8, title="Water yield 1981-2010") +
  scale_fill_brewer(name="Q mm", palette='RdYlBu', drop=FALSE)
```

## Cooks Partisan Voting Index

### Acquiring county level presidential data

To calculate county level PVI, we need to aggregate county level presidential election results for the years of interest. The steps to create the data file `presidential_votes_2004_2008.csv` is included below. Note, it requires the use of a subscription to CQ press.I used Vanderbilt Universities.

1. Go to http://library.cqpress.com.proxy.library.vanderbilt.edu/elections/download-data.php
2. Go to the main page >> Dowload data >> president >> county detail >> select state and year
3. The state, county, race date, total number of votes, percent democratic, and percent republican are extracted from each downloaded file and combined into the `presidential_votes_2004_2008.csv` file. Unfortunately, the county FIPS codes are not included with the CQpress downloads and must be added using the county name and state.

Below is the first 10 rows of the `presidential_votes_2004_2008.csv` file:

```{r, echo=F}
pres_data <- read.csv("data/presidential_votes_2004_2008.csv", stringsAsFactors=F)
```

```{r, echo=F, results='asis'}
pres_data.tab <- xtable(pres_data,digits=c(0,0,0,0,0,2,2))
align(pres_data.tab) <- rep("l",7)
print(pres_data.tab[1:10,],include.rownames=F,comment = F)
```

### Adding county FIPS codes

We want to use the county and state name to add the county fips codes from `wudata`. A good bit of pre-processing is needed prior to trying to add FIPS codes to the `pres_data` file. Some of the county names may have spaces, capital letters, or puncuation that is inconsistent and will cause an incorrect merger. There are also other possible differences (e.g., different spellings, abbreviations, or how a county is treated by the Census Bureau), which is addressed later.

```{r}
# prepare for merging with pres_data file
cnty.fips <- wudata %>%
  select(state,county,cnty_fips) %>%
  mutate(state = str_to_lower(state) %>%
           str_replace_all(fixed(" "), ""),
         county = str_to_lower(county) %>%
           str_replace_all("parish", "") %>%
           str_replace_all("[[:punct:]]","") %>%
           str_replace_all("[[:digit:]]","") %>%
           str_replace_all(fixed(" "), "") %>%
           ifelse(grepl("obrien",.), "obrien",.) %>%
           ifelse(grepl("princegeorges",.), "princegeorges",.) %>%
           ifelse(grepl("queenannes",.), "queenannes",.) %>%
           ifelse(grepl("stmarys",.), "stmarys",.) %>%
           ifelse(grepl("carsoncity",.),"carsoncity",.)%>%
           ifelse(grepl("stlouiscity",.),"stlouiscity",.)) %>%
  mutate(county = ifelse(state=="virginia",str_replace_all(county,"city", ""),county)) 

# two missing counties for 2008 election
# http://uselectionatlas.org
pres_2008_miss <- data.frame(cnty_fips=c("29001","45001"),
                             year=c(2008,2008),
                             state=c("missouri","southcarolina"),
                             total_votes=c(11871,11001),
                             rep_prop=c(0.4963,0.5694),
                             dem_prop=c(0.4831,0.4175),
                             stringsAsFactors = F)

# Load pres_data file, clean, and merge
pres_data <- read.csv("data/presidential_votes_2004_2008.csv", stringsAsFactors=F) %>%
  mutate(county = str_replace_all(county, "[[:punct:]]","") %>% # remove punctuation
           str_replace_all(fixed(" "), "") %>% # remove white spaces
           str_to_lower(),
         state = str_to_lower(state) %>% # make state name lower case
           str_replace_all(fixed(" "), "")) %>% 
  mutate(county = ifelse(state=="virginia",str_replace_all(county,"city", ""),county)) %>%
  filter(!county %in% c("kansascity","votesnotreportedbycounty")) %>% # remove flag
  filter(!(state %in% c("alaska","hawaii"))) %>%
  left_join(cnty.fips,by=c("county","state")) %>%
  group_by(cnty_fips, year, state) %>%
  dplyr::summarize(total_votes=sum(total_votes),
            rep_prop = mean(rep_prop),
            dem_prop = mean(dem_prop)) %>%
  ungroup() %>%
  bind_rows(pres_2008_miss) %>%
  arrange(cnty_fips) %>%
  as.data.frame()

```

### Calculate PVI and

The index involves comparing the two-way presidential vote for two consecutive years. This means we only care about head-to-head results for just the republican votes and democratic votes. I used 2004 and 2008. The variables are straight forward:

$DY1_n$ = Democratic share of the national two-way Presidential vote, year one (2004)

$DY2_n$ = Democratic share of the national two-way Presidential vote, year two (2008)

$DY1_c$ = Democratic share of the county two-way Presidential vote, year one (2004)

$DY1_c$ = Democratic share of the county two-way Presidential vote, year one (2008)


$$PVI = \left(\frac{DY1_d + DY2_d}{2} - \frac{DY1_n + DY2_n}{2}\right) * 100$$

Where "democratic share" is the $dem_{votes}/(dem_{votes} + rep_{votes})$. This could just as easily be calculated as percent republican. For above, a positive number would mean more democratic than the national average, and a negative number would mean more republican.

We are starting with two years of total votes and the proportion that are democratic or republican.

```{r}
# 2004 rep and dem head to head
pres2004 <- pres_data %>%
  filter(year==2004) %>%
  #filter(!is.na(cnty_fips)) %>%
  mutate(dem04 = dem_prop*total_votes, 
         rep04 = rep_prop*total_votes, 
         rep_dem04 = dem04 + rep04,
         dem2004 = dem04/rep_dem04) %>%
  select(cnty_fips,dem04,rep_dem04)

# 2008 rep and dem head to head
pres2008 <- pres_data %>%
  filter(year==2008) %>%
  #filter(!is.na(cnty_fips)) %>%
  mutate(dem08 = dem_prop*total_votes, 
         rep08 = rep_prop*total_votes, 
         rep_dem08 = dem08 + rep08,
         dem2008 = dem08/rep_dem08) %>%
  select(cnty_fips,dem08,rep_dem08)


# calculate national share
dn04 <- sum(pres2004$dem04)/sum(pres2004$rep_dem04)
dn08 <- sum(pres2008$dem08)/sum(pres2008$rep_dem08)
mn_nat_dem = (dn08 + dn04)/2

# calculate pvi for counties
cnty_pvi <- pres2004 %>%
  left_join(pres2008, by="cnty_fips") %>%
  mutate(pdem08 = dem08/rep_dem08,
         pdem04 = dem04/rep_dem04,
         mn_dem = (pdem08 + pdem04)/2,
         cnty.pvi = (mn_dem - mn_nat_dem)*100)

write.csv(cnty_pvi, file="cnty_pvi.csv", row.names=F)
```

```{r, echo=F}
pvi.map <- cnty_pvi %>%
  select(cnty_fips,cnty.pvi) %>%
  set_colnames(c("region","value")) %>%
  mutate(region=as.integer(region)) 

county_choropleth(pvi.map, num_colors = 1, title="County level PVI for 2008 and 2004") +
  scale_fill_gradient2(low = "red",
                       mid = "white",
                       high = "blue",
                       midpoint = 0,
                       na.value = "black",
                       breaks = pretty(pvi.map$value, n = 10))
```

```{r, echo=F}
## cooks partisan voting index
cnty_pvi <- read.csv("data/cnty_pvi.csv", stringsAsFactors = F) %>%
  select(cnty_fips,cnty.pvi) %>%
  mutate(cnty_fips = as.character(cnty_fips) %>%
           str_pad(5, pad = "0"))
```

## land area

```{r, eval=F}
# url for area data
area.url <- "http://www2.census.gov/prod2/statcomp/usac/excel/LND01.xls"

# download area data file
download.file(area.url, "data/cnty_area.xls")

# county land area in sq miles
cnty.area <- read_excel("data/cnty_area.xls") %>%
  select(cnty_fips=STCOU,cnty_area=LND110200D)
```

```{r, echo=F}
## county land area in sq miles
cnty.area <- read_excel("data/cnty_area.xls") %>%
  select(cnty_fips=STCOU,cnty_area=LND110200D)
```

## Census demographic data

This section uses the `acs` package and table numbers from the american community survey. Description can be found [here](https://www.socialexplorer.com/data/ACS2011_5yr/metadata/?ds=ACS11_5yr).

```{r, echo=F}
library(acs)
api.key.install(key="3143f097ff908575e23f4d7c054d7dc8612a14a6")
```

```{r}
# make geo object of counties
cnty.geo <- geo.make(state="*", county="*")

# lookup function 
#acs.lookup(keyword = "median income", endyear = 2010, span=5, case.sensitive=F)

# population growth 2000 to 2010
pop2000 <- acs.fetch(endyear=2000, geography = cnty.geo, dataset="sf1", variable="P001001")
pop2010 <- acs.fetch(endyear=2010, geography = cnty.geo, dataset="sf1", variable="P0010001")

cnty.pop2000 <- data.frame(geography(pop2000), estimate(pop2000)) %>%
  mutate(cnty_fips = str_pad(paste0(state,county),5,pad="0")) %>%
  select(cnty_fips, pop2000=P001001)

cnty.pop2010 <- data.frame(geography(pop2010), estimate(pop2010)) %>%
  mutate(cnty_fips = str_pad(paste0(state,county),5,pad="0")) %>%
  select(cnty_fips, pop2010=P0010001)

popgrowth <- left_join(cnty.pop2000,cnty.pop2010, by="cnty_fips") %>%
  mutate(pgrowth = round((pop2010-pop2000)/pop2000,3)) %>%
  select(cnty_fips, pgrowth)

# acs variable names and codes
acs.vars <- data.frame(var.names=c("cnty_pop","num_houses","med_income","gini",
                                   "poverty","renter","plumbing","medyr_structure",
                                   "med_age","college","sfh_d","sfh_a","mob_home",
                                   "apt10","apt20","apt50","med_value"),
                       var.codes=c("B01003_001","B25001_001","B06011_001","B19083_001",
                                   "B17001_002","B25003_003","B25047_002","B25035_001",
                                   "B01002_001","B06009_004","B25024_002","B25024_003",
                                   "B25024_010","B25024_007","B25024_008","B25024_009",
                                   "B25077_001"),
                       stringsAsFactors = F)

# fetch acs data for counties
acs.data.raw <- acs.fetch(endyear=2010, span=5, cnty.geo, variable=acs.vars$var.codes,
                          col.names=acs.vars$var.names)

# Build data.frame of census data
acs.data <- data.frame(geography(acs.data.raw), estimate(acs.data.raw)) %>%
  mutate(cnty_fips = str_pad(paste0(state,county),5,pad="0")) %>%
  left_join(cnty.area, by="cnty_fips") %>%
  left_join(popgrowth, by="cnty_fips") %>%
  na.omit() %>%
  mutate(ppl_house=cnty_pop/num_houses,
         house_dens=num_houses/cnty_area,
         prent=renter/num_houses,
         med_age_struc=2010-medyr_structure,
         pcollege=college/cnty_pop,
         ppoverty=poverty/cnty_pop,
         psfh=(sfh_a + sfh_d)/num_houses,
         pmob_home=mob_home/num_houses,
         papt=(apt10+apt20+apt50)/num_houses) %>%
  select(cnty_fips,cnty_pop:papt) %>%
  set_rownames(NULL)
```

# Merge data sets

Create a list of the data sets and then merge everything by county FIPS codes. Also calculate houshold water use in this step:

```{r}
data.list <- list(wudata,
                  cnty_typology,
                  wyield,
                  tmax05_10,
                  precip05_10,
                  cnty_pvi,
                  acs.data)

data.full <- Reduce(function(x,y) 
  left_join(x,y,by="cnty_fips"), data.list) %>%
  left_join(climate_region, by="state.abb") %>%
  mutate(house_served = (pop_served/cnty_pop.x) * num_houses,
         wh = ((wthdrw_tw*1e6*365))/house_served) 
```

## Water trading

Counties trade water with other counties. If a county is selling water, then that county will have exceptionally high household water use, and if a county is buying water, the opposite it will be true. Unfortunately, there is not yet a database that captures water trading information and our regression model cannot account for it's effects (model coefficients would be driven by the extremes due to water trading). In order to account for this, we used data from [Rockaway 2011](https://www.researchgate.net/publication/279937580_Residential_Water_Use_Trends_in_North_America) (which draws on data from an earlier [EPA report](http://www.waterrf.org/PublicReportLibrary/4031.pdf)) to constrain our data set to counties that fall within a reasonable range of values that were taken from the above linked paper. The contraints are based off the AWWA water survey, and provides thresholds from a source that is independent of the USGS data. We dropped counties that with annual household water use values less than 50,000 gal/yr and greater than 300,000 gal/yr. 

```{r, echo=F, fig.height=4}
ggplot(data.full, aes(wh)) + 
  geom_histogram(bins="100",color="white") + theme_bw() +
  scale_x_log10(breaks=c(1e4,1e5,1e6), labels=c("10,000","100,000","1,000,000")) + 
  geom_vline(xintercept=102985, linetype="dashed",color="red", size=1) +
  geom_vline(xintercept=5e4, linetype="dashed",color="blue", size=1) +
  geom_vline(xintercept=3e5, linetype="dashed",color="blue", size=1) +
  labs(x="withdrawal/households (gal/house/year)") +
  coord_cartesian(xlim=c(1e3,1e7)) +
  ggtitle("Thresholds derived from Rockaway et al., 2011 paper") +
  labs(subtitle="red line = mean from table 3 in paper, blue lines = 50,000 and 300,000 gal/year")
```


Select only the variables on interest, drop counties outside of thresholds detailed above, and scale the covariates (i.e., convert to z-scores).

```{r}
model.data <- data.full %>%
  select(cnty_fips,county,state,state.abb,climate_region, 
         rur2urb,econdep,retire,wh,cnty_pop=cnty_pop.y,
         pgrowth,prop_sw,Qmm,tmax_40,tmax_0510,tmax_diff,
         ppt_40,ppt_0510,ppt_diff,med_income,ppoverty,
         gini,med_age,pcollege,cnty.pvi,house_dens,
         ppl_house,med_value,med_age_struc,prent,
         psfh,pmob_home,papt) %>%
  na.omit() %>%
  filter(wh > 5e4 & wh < 3e5) %>%
  mutate_each(funs(as.numeric(scale(.))), -(cnty_fips:wh))
```

We dropped 563 counties which leaves us with 2546 for the analysis. Below is a map of the remaining counties:

```{r, echo=F}
wh.map <- model.data %>%
  select(region=cnty_fips,value=wh) %>%
  mutate(region=as.integer(region))

county_choropleth(wh.map, num_colors = 8, title="Houshold withdrawals gal/yr") 
```

See which covariates have a correlation greater than 0.4. This is a completely arbitrary cutoff:

```{r}
high.cor <- model.data %>%
  select(wh:papt) %>%
  correlate() %>%
  shave() %>%
  stretch() %>%
  na.omit() %>%
  filter(abs(r)>0.4)
```

\newpage

```{r, echo=F, results='asis'}
cor.tab <- xtable(high.cor,digits=c(0,0,0,3))
align(cor.tab) <- c("l","l","l","c")
print(cor.tab,include.rownames=F,comment = F)
```

Finally, remove several variables that are highly correlated with other variables:

```{r}
model.data <- model.data %>%
  select(-c(tmax_0510,ppt_0510,ppoverty,
            med_value,pmob_home))
```



